---
title: "ggplot2 demo"
output: html_document
author: "EpiRHandbook Team"
date: "20 August 2021"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```


# Introduction  

This R markdown script contains the R code used in the ggplot2 / data visualization demonstration.  
The intention is that you can run these codes at your leisure and explore how the syntax works.  

* Each code chunk below corresponds to a topic covered in the accompanying slide deck.  
* Note:  some of the early slide deck plots had their text size enlarged with `theme_grey(base_size = 16)` to ease readability on the slides. This modification is not added in this script, until the standard theme commands are introduced.  


This script includes the R code to produce the **ggplot2** figures in these [presentation slides](https://appliedepi.github.io/emory_training/presentation/slides_ggplot.html) 

# Load packages

```{r, warning=F, message=F}
pacman::p_load(
  rio,            # import/export
  here,           # file locator
  DT,             # interactive tables for this report
  lubridate,      # working with dates
  incidence2,     # epidemic curves
  apyramid,       # demographic pyramids
  gghighlight,    # specialized highlight plots
  ggrepel,        # smart labels
  scales,         # formatting of scales
  cowplot,        # combining plots
  ggplotly,       # interactive plots
  tidyverse       # mega-package for data management and visualization
  )
```


# Import cleaned linelist data

We assume that you are running this script from within the downloadable (zipped folder) R project called "case_study".  
The directory structure assumed by the below command is that the "linelist_cleaned.rds" dataset is stored within at the file path "data/covid_example_data/linelist_cleaned.rds", from the R project root folder.  

```{r}
linelist <- import(here::here("data", "covid_example_data", "linelist_cleaned.rds"))
```


# Review first rows of data

```{r}
head(linelist, 25) %>% 
  DT::datatable(
    options = list(
      scrollY = 300,
      scrollX = 600,
      pageLength = 25,
      fontSize = '25%',
      dom = "ti",
      ordering = FALSE,
      rownames = FALSE,
      options = list(pageLength = 5, scrollX=T),
      class = 'white-space: nowrap' 
    )
  )
```



# Empty ggplot  

```{r}
ggplot(data = linelist)
```

# Add axis mappings  

```{r}
ggplot(data = linelist,
       mapping = aes(
         x = age,
         y = days_hosp))
  
```


# Add geometry  

```{r}
ggplot(
  data = linelist,
  mapping = aes(
    x = age,
    y = days_hosp))+
geom_point()
```

# Add further geometry

```{r}
ggplot(
  data = linelist,
  mapping = aes(
    x = age,
    y = days_hosp)) +
geom_point() +
geom_smooth()
```


# Static aesthetic

```{r}
ggplot(
  data = linelist,
  mapping = aes(
     x = age,
     y = days_hosp)) +
geom_point(color = "seagreen")
```

# Dynamic aesthetics  

```{r}
ggplot(
  data = linelist,
  mapping = aes(
    x = age,
    y = days_hosp,
    color = died)) +   
geom_point()
```


# Static and dynamic aesthetics

```{r}
ggplot(
  data = linelist,
  mapping = aes(
    x = age,
    y = days_hosp,
    color = died))+ # color varies by outcome
geom_point(size = 7,    # all points 7x size
           alpha = 0.7) # transparency    
```


# A common error:  

```{r}
# create data
outcome_time_data <- linelist %>% 
  group_by(died,
           week_onset = lubridate::floor_date(date_onset, "week")) %>% 
  summarise(n_cases = n()) %>%

  select(week_onset, everything()) %>% 
  arrange(week_onset) %>% 
  drop_na(week_onset)

# Show top
outcome_time_data %>%
  head() %>% 
  knitr::kable()

# plot with error
ggplot(
  data = outcome_time_data,
  mapping = aes(
    x = week_onset,       
    y = n_cases)) +
  geom_line() +
  labs(title = "Weekly cases, by outcome (?)")

# plot resolved
ggplot(
  data = outcome_time_data,
  mapping = aes(
    x = week_onset,       
    y = n_cases,
    color = died)) +
  geom_line() +
  labs(title = "Weekly cases, by outcome")

```



# Facets/small multiples

```{r, fig.width=10, fig.height=10}

# facet wrap by ethnicity/race value
ggplot(data = linelist,
       mapping = aes(x = date_onset)) +
  geom_histogram() +
  facet_wrap(~ eth_race)

# facet wrap with free y-axes
ggplot(data = linelist,
       mapping = aes(x = date_onset)) +
  geom_histogram() +
  facet_wrap(~ eth_race, scales = "free_y") +
  labs(caption = "Note the free y-axis")

# facet wrap with two variables
ggplot(
  data = linelist,
  mapping = aes(x = date_onset)) +
  geom_histogram() +
  facet_wrap(gender ~ eth_race)

# facet grid with two variables
ggplot(
  data = linelist,
  mapping = aes(x = date_onset)) +
  geom_histogram() +
  facet_grid(gender ~ eth_race)


# drop levels you do not want
linelist %>% 
  drop_na(gender, eth_race) %>% 
  filter(gender != "Unknown") %>% 
  
  ggplot(
    mapping = aes(x = date_onset)) +
  geom_histogram() +
  facet_grid(gender ~ eth_race)




```



# gghighlight

```{r}

# facets with gghighlight and fill
ggplot(
  data = linelist,
  mapping = aes(
    x = date_onset,
    fill = eth_race)) +
  geom_histogram() +
  facet_wrap(~ eth_race) +
  gghighlight::gghighlight()


# geom_line with gghighlight
linelist %>% 
  # get daily counts by zip code
  group_by(age_group, date_report) %>%
  count() %>% 
  
  # plot
  ggplot(
    mapping = aes(
      x = date_report,
      y = n,
      color = age_group)) +
  geom_line()+
  gghighlight::gghighlight(age_group %in% c("40-49", "60-69")) +
  theme(legend.position = "none")
```


# Scales - default

```{r}
ggplot(linelist) +
  geom_bar(aes(x = eth_race, fill = died)) +
  theme_gray(base_size = 18)
```


# Scales - adjusted fill  



```{r}
ggplot(linelist) +
geom_bar(aes(x = eth_race, fill = died)) +
scale_fill_manual(        
  values = c("Yes" = "violetred", 
             "No" = "aquamarine",
            "Unknown" = "grey")) +
theme_gray(base_size = 18)
```

Note that you can adjust the `fig.width=` and `fig.height=` parameters of the R Markdown chunk to make the image larger or smaller.
This can resolve the problem of text being too large.  

```{r, fig.width=12, fig.height=8}
ggplot(linelist) +
geom_bar(aes(x = eth_race, fill = died)) +
scale_fill_manual(        
  values = c("Yes" = "violetred", 
             "No" = "aquamarine",
            "Unknown" = "grey")) +
theme_gray(base_size = 18)
```

# Scales - adjusted y-axis  

```{r}
ggplot(linelist) +
geom_bar(aes(x = eth_race, fill = died)) +
scale_fill_manual(        
  values = c("Yes" = "violetred", 
             "No" = "aquamarine",
            "Unknown" = "grey")) +
scale_y_continuous(
  breaks = seq(from = 0,
               to = 35000,
               by = 5000)) +
theme_gray(base_size = 18)
```

# Scales - adjusted expansion

```{r}
ggplot(linelist) +
geom_bar(aes(x = eth_race, fill = died))+
scale_fill_manual(        
  values = c("Yes" = "violetred", 
             "No" = "aquamarine",
            "Unknown" = "grey")) +
scale_y_continuous(
  breaks = seq(from = 0,
               to = 35000,
               by = 5000),
  expand = c(0,0)) +
scale_x_discrete(
  expand = c(0,0)) +
theme_grey(base_size = 18)
```

# Scales - date axis

```{r}
ggplot(linelist) +
geom_histogram(aes(x = date_onset)) +
scale_x_date(
  date_breaks = "3 months",
  date_labels = "%d %b\n%Y"
) +
theme_grey(base_size = 18)

```

