---
title: "ggplot2 demo"
output: html_document
author: "EpiRHandbook Team"
date: "20 August 2021"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```


# Introduction  

This R markdown script contains the R code used in the ggplot2 / data visualization demonstration.  
The intention is that you can run these codes at your leisure and explore how the syntax works.  

* Each code chunk below corresponds to a topic covered in the accompanying slide deck.  
* Note:  some of the early slide deck plots had their text size enlarged with `theme_grey(base_size = 16)` to ease readability on the slides. This modification is not added in this script, until the standard theme commands are introduced.  


This script includes the R code to produce the **ggplot2** figures in these [presentation slides](https://appliedepi.github.io/emory_training/presentation/slides_ggplot.html) 

# Load packages


```{r, warning=F, message=F}
## install and load necessary packages 
pacman::p_load(
  rio,        # importing data  
  here,       # relative file pathways  
  skimr,      # review data
  DT,         # visualize data frame
  janitor,    # data cleaning and tables
  epikit,     # age categories
  lubridate,  # working with dates
  incidence2, # epidemic curves
  ggrepel,    # smart labels
  ggExtra,    # extras
  esquisse,   # point-and-click for simple ggplots
  apyramid,   # age pyramids
  scales,     # formatting of scales
  plotly,     # interactive plots
  cowplot,    # combine plots
  tidyverse  # ggplot2 & data management
)

```


# Import cleaned linelist data

We assume that you are running this script from within the downloadable (zipped folder) R project called "case_study".  
The directory structure assumed by the below command is that the "linelist_cleaned.rds" dataset is stored within at the file path "data/covid_example_data/linelist_cleaned.rds", from the R project root folder.  

```{r}
linelist <- import(here::here("data", "covid_example_data", "linelist_cleaned.rds"))
```


# Review first rows of data

```{r}
head(linelist, 25) %>% 
  DT::datatable(
    options = list(
      scrollY = 300,
      scrollX = 600,
      pageLength = 25,
      fontSize = '25%',
      dom = "ti",
      ordering = FALSE,
      rownames = FALSE,
      options = list(pageLength = 5, scrollX=T),
      class = 'white-space: nowrap' 
    )
  )
```



# Empty ggplot  

```{r}
ggplot(data = linelist)
```

# Add axis mappings  

```{r}
ggplot(
  data = linelist,
  mapping = aes(
   x = age,
   y = days_hosp))
```


# Add geometry  

```{r}
ggplot(
  data = linelist,
  mapping = aes(
   x = age,
   y = days_hosp)) +
geom_point()
```


# Sample geoms  

```{r}
# Histogram
histogram <- ggplot(
    data = linelist,
    mapping = aes(
      x = date_onset))+
  geom_histogram(fill = "darkorange")+
  labs(title = "geom_histogram()")+
  theme_minimal(base_size = 16)
  
histogram

# Points
points <- ggplot(
    data = linelist,
    mapping = aes(
      x = age,       
      y = days_hosp)
    )+
  geom_point()+
  labs(title = "geom_point()")+
  theme_minimal(base_size = 16)

points

# Lines
lines <- linelist %>% 
  group_by(eth_race,
           week_report = lubridate::floor_date(date_report, "week")) %>% 
  summarise(n_cases = n()) %>% 
  drop_na(week_report) %>% 
  ggplot(
    mapping = aes(
      x = week_report,       
      y = n_cases,
      color = eth_race),
    size = 4
    )+
  geom_line()+
  labs(title = "geom_line()")+
  theme_minimal(base_size = 16)

lines

# Bars
bars <- linelist %>% 
  ggplot(mapping = aes(
      y = died,
      fill = died))+
  geom_bar()+
  labs(title = "geom_bar()")+
  theme_minimal(base_size = 16)+
  theme(legend.position = "none")

bars

# Boxplots
boxplots <- ggplot(
    data = linelist,
    mapping = aes(
      x = died,       
      y = age)
    )+
  geom_boxplot()+
  labs(title = "geom_boxplot()")+
  theme_minimal(base_size = 16)

boxplots

# Violin plots
violins <- ggplot(
    data = linelist,
    mapping = aes(
      x = died,       
      y = days_hosp,
      fill = died)
    )+
  geom_violin()+
  labs(title = "geom_violin()")+
  theme_minimal(base_size = 16)

violins
```



# Add further geometry

```{r}
ggplot(
  data = linelist,
  mapping = aes(
    x = age,
    y = days_hosp)) +
geom_point() +
geom_smooth()
```


# Static aesthetic

```{r}
ggplot(
  data = linelist,
  mapping = aes(
     x = age,
     y = days_hosp)) +
geom_point(color = "seagreen")
```

# Dynamic aesthetics  

```{r}
ggplot(
  data = linelist,
  mapping = aes(
    x = age,
    y = days_hosp,
    color = died)) +   
geom_point()
```


# Static and dynamic aesthetics

```{r}
ggplot(
  data = linelist,
  mapping = aes(
    x = age,
    y = days_hosp,
    color = died))+      # color varies by outcome
geom_point(size = 7,     # all points 7x size
           alpha = 0.7)  # transparency  
```


# Aesthetic mapping placement

The below example demonstrates how certain aesthetics are inherited by subsequent geoms (e.g. x and y axes), while others are not (e.g. size = 1). The final `geom_smooth()` command utilized the entire dataset, whereas the previous one produces one curve per `died` value.  


```{r}
# begin plot, assign x and y axes
ggplot(                
  data = linelist,      
  mapping = aes(        
    x = age,
    y = days_hosp)) +

# points colored by outcome
geom_point(
  mapping = aes(color = died),
  size = 1) +

# smoothed means by outcome
geom_smooth(
  mapping = aes(color = died),
  size = 3) +

# smoothed mean of entire dataset
geom_smooth(color = "black")
```


# A common error:  
```{r, eval=T, echo = T, warning=F, message=F}
# aggregate to counts dataset
outcome_week_data <- linelist %>% 
  count(
    died,
    week_onset = lubridate::floor_date(
      x = date_onset,
      unit = "week")) %>% 

  select(week_onset, everything()) %>% 
  arrange(week_onset) %>% 
  drop_na(week_onset)

# display first 8 rows
outcome_week_data %>%
  head(8) %>% 
  knitr::kable()
```

```{r}
# plot with error
ggplot(
  data = outcome_week_data,
  mapping = aes(
    x = week_onset,       
    y = n)) +
geom_line()

# plot resolved
ggplot(
  data = outcome_week_data,
  mapping = aes(
    x = week_onset,       
    y = n,
    color = died)) +
geom_line()

```



# Facets/small multiples

```{r, fig.width=10, fig.height=10}

# facet wrap by ethnicity/race value
ggplot(
  data = linelist,
  mapping = aes(x = date_onset)) +
geom_histogram() +
facet_wrap(~ eth_race)

# facet wrap with free y-axes
ggplot(
  data = linelist,
  mapping = aes(x = date_onset)) +
geom_histogram() +
facet_wrap(~eth_race,
           scales = "free_y") +
labs(caption = "Note the free y-axis")

# facet wrap with two variables
ggplot(
  data = linelist,
  mapping = aes(x = date_onset)) +
geom_histogram() +
facet_wrap(eth_race ~ gender)

# facet grid with two variables
ggplot(
  data = linelist,
  mapping = aes(x = date_onset)) +
geom_histogram() +
facet_grid(eth_race ~ gender)


# drop levels you do not want
linelist %>% 
  drop_na(gender, eth_race) %>% 
  filter(gender != "Unknown") %>% 
  
  ggplot(
    mapping = aes(x = date_onset)) +
  geom_histogram() +
  facet_grid(eth_race ~ gender)




```



# gghighlight

```{r}

# facets with gghighlight and fill
ggplot(
  data = linelist,
  mapping = aes(
    x = date_onset,
    fill = eth_race)) +
geom_histogram() +
facet_wrap(~ eth_race) +
gghighlight::gghighlight()


# geom_line with gghighlight
linelist %>% 
  # get daily counts by zip code
  group_by(age_group, date_report) %>%
  count() %>% 
  
  # plot
  ggplot(
    mapping = aes(
      x = date_report,
      y = n,
      color = age_group)) +
  geom_line()+
  gghighlight::gghighlight(age_group %in% c("40-49", "60-69")) +
  theme(legend.position = "none")
```


# Scales - default

```{r}
ggplot(
  data = linelist,
  mapping = aes(
    x = eth_race,
    fill = died)) +
geom_bar() 
```


# Scales - adjusted fill  



```{r}
ggplot(
  data = linelist,
  mapping = aes(
    x = eth_race,
    fill = died)) +
geom_bar() +
scale_fill_manual(        
  values = c(
    "Yes" = "violetred", 
    "No" = "aquamarine",
    "Unknown" = "grey"))
```


# Scales - adjusted y-axis  

```{r}
ggplot(
  data = linelist,
  mapping = aes(
    x = eth_race,
    fill = died)) +
geom_bar() +
scale_fill_manual(        
  values = c("Yes"    = "violetred", 
             "No"     = "aquamarine",
            "Unknown" = "grey")) +
scale_y_continuous(
  breaks = seq(from = 0,
               to = 35000,
               by = 5000))
```

# Scales - Start axes at 0 

```{r}
ggplot(
  data = linelist,
  mapping = aes(
    x = eth_race,
    fill = died)) +
geom_bar() +
scale_fill_manual(        
  values = c("Yes" = "violetred", 
             "No" = "aquamarine",
            "Unknown" = "grey")) +
scale_y_continuous(
  breaks = seq(from = 0,
               to = 35000,
               by = 5000),
  expand = c(0,0)) +
scale_x_discrete(
  expand = c(0,0)) 
```

# Scales - date axis - default

```{r}
ggplot(
  data = linelist,
  mapping = aes(x = date_onset)) +
geom_histogram()

```

# Scales - date label breaks  

```{r}
ggplot(
  data = linelist,
  mapping = aes(x = date_onset)) +
geom_histogram() +
scale_x_date(
  date_breaks = "3 months")

```

# Scales - date axis labels
```{r}
ggplot(
  data = linelist,
  mapping = aes(x = date_onset)) +
geom_histogram() +
scale_x_date(
  date_breaks = "3 months",
  date_labels = "%d %b\n%Y")

```

# Scales - date axis labels

```{r}
ggplot(
  data = linelist,
  mapping = aes(x = date_onset)) +
geom_histogram() +
scale_x_date(
  date_breaks = "3 months",
  labels = scales::label_date_short())

```

# Scales - display proportions as percents  

We make a dataset with proportions  

```{r}
# no lag accounted for, for simplicity
CFR_data <- linelist %>% 
  group_by(month = floor_date(date_report, "month")) %>% 
  summarise(
    cases = n(),
    deaths = sum(died_covid == "Yes", na.rm=T),
    CFR = deaths / cases)

# display first few rows
CFR_data %>% 
  head(4) %>%
  knitr::kable()

```


```{r}
ggplot(
  data = CFR_data,
  mapping = aes(
    x = month,
    y = CFR)) +
geom_line(size = 2, color = "brown")+
scale_y_continuous(labels = percent)
```



# Plot labels  

```{r}
ggplot(data = linelist)+
geom_point(
  mapping = aes(
    x = age,
    y = days_hosp,
    color = died),    
  alpha = 0.3) +
labs(
  title = "Duration of admission",
  x = "Age (years)",
  y = "Duration (days)",
  caption = "Fictional COVID-19 data",
  color = "Deceased"
)
```


# Introducing dynamic labels

**Dynamic labels** - Imbed code in `str_glue()` that updates with the data  

```{r, eval=T, echo=T}
str_glue("Data as of {Sys.Date()}")
```

```{r, eval=T, echo=T}
str_glue("{fmt_count(linelist, is.na(date_onset))} cases missing onset and not shown")
```


# Standard themes


```{r, eval = TRUE, tidy = FALSE}

minimal <- ggplot(data = linelist,
       mapping = aes(
         x = age,
         y = days_hosp),    
       alpha = 0.3)+
  geom_point()+
  labs(
    title = "theme_minimal()",
    x = "Age (years)",
    y = "Duration (days)")+
  theme_minimal(base_size = 18)

minimal

bw <- ggplot(data = linelist,
       mapping = aes(
         x = age,
         y = days_hosp),    
       alpha = 0.3)+
  geom_point()+
  labs(
    title = "theme_bw()",
    x = "Age (years)",
    y = "Duration (days)")+
  theme_bw(base_size = 18)

bw

grey <- ggplot(data = linelist,
       mapping = aes(
         x = age,
         y = days_hosp),    
       alpha = 0.3)+
  geom_point()+
  labs(
    title = "theme_grey()",
    x = "Age (years)",
    y = "Duration (days)")+
  theme_grey(base_size = 18)

grey

dark <- ggplot(data = linelist,
       mapping = aes(
         x = age,
         y = days_hosp),    
       alpha = 0.3)+
  geom_point()+
  labs(
    title = "theme_dark()",
    x = "Age (years)",
    y = "Duration (days)")+
  theme_dark(base_size = 18)

dark

void <- ggplot(data = linelist,
       mapping = aes(
         x = age,
         y = days_hosp),    
       alpha = 0.3)+
  geom_point()+
  labs(
    title = "theme_void()",
    x = "Age (years)",
    y = "Duration (days)")+
  theme_void(base_size = 18)

void

light <- ggplot(data = linelist,
       mapping = aes(
         x = age,
         y = days_hosp),    
       alpha = 0.3)+
  geom_point()+
  labs(
    title = "theme_light()",
    x = "Age (years)",
    y = "Duration (days)")+
  theme_light(base_size = 18)

light

```



# theme() micro-adjustments  

```{r}
ggplot(
  data = linelist,
  mapping = aes(
     x = age,
     y = days_hosp,
     color = died),    
  alpha = 0.3)+
geom_point()+
labs(
  title = "Duration of admission",
  x = "Age (years)",
  y = "Duration (days)",
  color = "Deceased")+
theme_minimal(base_size = 16)+
theme(
  legend.position = "bottom",
  plot.title = element_text(color = "red",
                            size = 20,
                            face = "bold"),
  axis.title.y = element_text(angle = 0))
```




# geom_bar()

```{r}
ggplot(
  data = linelist,
  mapping = aes(
    x = eth_race)) +             
geom_bar() +
theme(axis.text.x= element_text(angle=30)) 
```

# geom_bar() with grouped rows  

```{r}
ggplot(
  data = linelist,
  mapping = aes(
      x = eth_race,   
      fill = died)) +   # stack bars by column "died"            
geom_bar()+    
theme_grey(base_size = 16)+
theme(axis.text.x= element_text(angle = 30))
```


# geom_col()

```{r}
# counts by race/ethnicity
linelist_eth <- linelist %>% 
  count(eth_race)
```

```{r}
ggplot(linelist_eth) +             
geom_col(           
    mapping = aes(   
      x = eth_race,    
      y = n))+
theme(axis.text.x= element_text(angle = 30))
```


# Bar plots common error

```{r}
linelist %>% 
  count(eth_race) %>% 
  ggplot()+
  geom_bar(
    mapping = aes(x = eth_race))
```


# geom_col() stacked

```{r}
# counts by outcome and race/ethnicity
linelist_eth_died <- linelist %>% 
  count(eth_race, died)
```

```{r}
ggplot(linelist_eth_died) + # begin with data aggregated by day/ethnicity              
  geom_col(                 # column height by aggregated counts
    mapping = aes(
      x = eth_race,         # one bar per ethnicity. 
      y = n,                # bar height from column "n"
      fill = died)) +       # stacked bars by outcome
theme_grey(base_size = 16)+
theme(axis.text.x= element_text(angle = 30))
```


# Flip axes

```{r}
ggplot(linelist) +             
  geom_bar(              
    mapping = aes(
      x = eth_race,          
      fill = died)) + 
coord_flip() + 
theme_grey(base_size = 16)+
theme(legend.position = "top") 
```


# Adjust bar order

```{r}
ggplot(linelist) +             
geom_bar(              
  mapping = aes(
    x = fct_infreq(eth_race),   
    fill = fct_infreq(died))) +  
coord_flip() + 
theme_grey(base_size = 16)+
theme(legend.position = "top") 
```




# Reverse bar order

```{r}
ggplot(linelist) +             
geom_bar(              # column height by number of rows
  mapping = aes(
    x = fct_rev(fct_infreq(eth_race)),            
    fill = fct_rev(fct_infreq(died)))) +     
coord_flip() + 
theme_grey(base_size = 16) +
theme(legend.position = "top") 
```


# Adjust bar width  

```{r}
ggplot(
  data = linelist_eth_died,
  mapping = aes(x = eth_race, y = n, fill = died)) +             
geom_col(width = 0.5) + 
theme_grey(base_size = 16)+
theme(axis.text.x = element_text(angle = 30))
```



# Adjacent bar plots

```{r}
ggplot(
  data = linelist_eth_died,
  mapping = aes(
    x = eth_race,
    y = n,
    fill = fct_infreq(died),
    label = n)) +             
geom_col(position = "dodge") + 
theme_grey(base_size = 16)+
theme(axis.text.x = element_text(angle = 30))
```


# Bar plots - display counts

```{r}
ggplot(linelist_eth_died,
       aes(x = eth_race, y = n, fill = died, label = n)) +             
geom_col() +
geom_text(size = 3, position = position_stack(vjust = 0.5)) + 
theme_grey(base_size = 16) +
theme(axis.text.x = element_text(angle=30))
```


# incidence2 - epicurves

```{r}
library(incidence2)

weekly <- incidence(
  x = linelist,            # define dataset
  date_index = date_onset, # date column
  interval = "week")       # bin interval   

plot(weekly)
```

# incidence2 - adjust interval

```{r}
bimonthly <- incidence(
  x = linelist,            
  date_index = date_onset, 
  interval = "2 months")    

plot(bimonthly)
```


# epicurve groups  

```{r}
weekly <- incidence(
  x = linelist,              
  date_index = date_onset,  
  interval = "weeks", 
  groups = eth_race) 

plot(weekly,
  fill = eth_race)+
  theme_minimal(base_size = 16)
```


# epicurves - add ggplot2 commands

```{r}
plot(weekly, fill = eth_race) +
scale_y_continuous(
  expand = c(0,0),
  breaks = seq(0,2000,250)) + 
scale_y_continuous(
 expand = c(0,0),
 breaks = seq(0,2000,250)) + 
theme_minimal(base_size = 16) + 
theme(legend.position = "top") +
labs(fill = "Race and\nEthnicity")
```

# epicurves - date axis

```{r}
plot(weekly,
  fill = eth_race,
  date_format = "%a %d %b %Y\n (Week %W)", 
  angle = 30)+                          
scale_y_continuous(
  expand = c(0,0),
  breaks = seq(0,2000,250)) + 
theme_minimal(base_size = 16) + 
theme(legend.position = "top") +
labs(fill = "Race and\nEthnicity")
```


# epicurves - show cases

Make small dataset to demonstrate with.  

```{r}
small_outbreak <- linelist %>% 
  filter(zip == 30024,
         date_report >= as.Date("2020-12-01")) %>% 
  incidence(
    # dataset is piped in from above
    date_index = date_onset,               
    interval = "Sunday weeks",             
    groups = eth_race)
```

```{r}
plot(small_outbreak,
  fill = eth_race,
 show_cases = TRUE)+
theme(legend.position = "bottom")+
labs(
  title = "ZIP 30024 by race") +
theme_minimal(base_size = 16)
```

# Epicurves - color palettes

```{r}
plot(weekly, fill = eth_race)+
   scale_fill_viridis_d(
    name = "Age Category",  # legend name
    na.value = "grey")      # for missing values
```


# epicurves from counts

```{r}
# For demo: aggregate linelist
linelist_day_counts <- linelist %>% 
  count(
    day = floor_date(date_report, "day"),
    died) %>% 
  drop_na(day)
```

```{r}
died_curve <- incidence(   
  linelist_day_counts,   # dataset with counts aggregated by day
  date_index = day,      # column with dates
  count = n,             # column with counts
  interval = "week",     # aggregate daily counts up to weeks
  groups = died          # group by outcome
  )
```

```{r}
# plot the weekly incidence epi curve, with stacked bars by hospital
plot(died_curve,         # incidence object
     fill = died)        # color the bars by outcome
```


# Demographic pyramids

Use the **apyramid** package.  

Clean the linelist to only two gender values

```{r}
linelist <- linelist %>% 
  mutate(
    gender = na_if(gender, "Unknown"), # convert to NA 
    gender = fct_drop(gender))         # drop unused factor levels
```

```{r}
age_pyramid(
  data = linelist,
  age_group = "age_group",
  split_by = "gender")
```

# Pyramids - options  

```{r}
age_pyramid(
  data = linelist,
  age_group = "age_group",
  split_by = "gender",
  proportional = TRUE,
  show_midpoint = FALSE,
  pal = c("darkgreen", "brown"))+
theme_minimal(base_size = 10)
```


# Pyramids - stacked

```{r}
age_pyramid(
  data = linelist,
  age_group = "age_group",
  split_by = "gender",
  stack_by = "hospitalized",
  proportional = TRUE,
  show_midpoint = FALSE,
  pal = c("darkgreen", "brown",
          "yellow", "orange"))+
theme_minimal(base_size = 10)

```


# Pyramids - add ggplot2 commands

```{r}
age_pyramid(
  data = linelist,
  age_group = "age_group",
  split_by = "gender",
  proportional = TRUE,
  show_midpoint = FALSE) +
theme_minimal(base_size = 10) +
labs(
  title = "Age and Gender",
  subtitle = "Fulton County, GA",
  x = "Percent of total",
  y = "Age group",
  fill = "Gender",
  caption = "Caption here")
```


# Complex dynamic caption

```{r, eval=T, echo=T}
str_glue("{missing} missing age or gender not shown.",
        missing = fmt_count(linelist, is.na(gender) | is.na(age_group))
        )
```

# Dynamic subtitle with date formatting

```{r, eval=T, echo=T}
str_glue("Fulton County, reported {min_date} - {max_date}",
    min_date = format(min(linelist$date_report, na.rm=T), "%B %d %Y"),
    max_date = format(max(linelist$date_report, na.rm=T), "%B %d %Y")
    )
```

# Pyramids with dynamic labels

```{r, eval=T, echo=T}
age_pyramid(
  data = linelist,
  age_group = "age_group",
  split_by = "gender",
  proportional = TRUE,
  show_midpoint = FALSE)+
theme_minimal(base_size = 16)+
labs(
  title = "Age and Gender",
  subtitle = str_glue(
  "Fulton County, reported {min_date} - {max_date}",
  
  min_date = format(
    min(linelist$date_report, na.rm=T),
    "%b %d %Y"),
  
  max_date = format(
    max(linelist$date_report, na.rm=T),
    "%b %d %Y")
  ),
  x = "Percent of total",
  y = "Age group",
  fill = "Gender",
  caption = str_glue(
    "{missing} missing age or gender not shown.",
    missing = fmt_count(
      linelist,
      is.na(gender) | is.na(age_group))
    )
  )
```


# Labeling  

Make CFR dataset

```{r}
race_CFR_age <- linelist %>% 
  group_by(eth_race) %>% 
  summarise(
    cases = n(),
    deaths = sum(died_covid == "Yes",
                 na.rm=T),
    CFR = deaths/cases,
    med_age = median(age, na.rm=T),
    med_dur = median(days_hosp, na.rm=T)
  )

```

```{r, echo=T, eval=T}
head(race_CFR_age, 5) %>% 
  knitr::kable()
```


# Simple labels with geom_text()

```{r}
ggplot(
  data = race_CFR_age,
  mapping = aes(
    x = med_age,
    y = CFR,
    size = cases,
    label = eth_race)) +
  geom_point() +
  geom_text()
```


# Better labels with geom_label_repel()

```{r}
ggplot(
  data = race_CFR_age,
  mapping = aes(
    x = med_age,
    y = CFR,
    size = cases,
    label = eth_race)) +
  geom_point() +
 geom_label_repel(
    size = 5,
    min.segment.length = 0)
```

# Improve labels with dynamic display

```{r}
ggplot(
  data = race_CFR_age,
  mapping = aes(
    x = med_age,
    y = CFR,
    size = cases,
    label = str_glue(
      "{eth_race}\n{cases} cases"))) +
  geom_point() +
 geom_label_repel(
   size = 5,
   min.segment.length = 0)
```


# Improve labels with comma display 

```{r}
ggplot(
  data = race_CFR_age,
  mapping = aes(
    x = med_age,
    y = CFR,
    size = cases,
    label = str_glue(
      "{eth_race}\n{comma(cases)} cases"))) +
geom_point() +
geom_label_repel(
   size = 4,
   min.segment.length = 0)
```


# Add marginal distributions to scatterplot

```{r}
scatterplot <- ggplot(
  data = linelist,
  mapping = aes(
    x = age,
    y = days_hosp)) + 
geom_point()

ggExtra::ggMarginal(
  scatterplot,                   # add marginal histograms
  type = "histogram",            # specify histograms
  fill = "lightblue",            # bar fill
  xparams = list(binwidth = 10), # other parameters for x-axis marginal
  yparams = list(binwidth = 5))  # other parameters for y-axis marginal
```


# Combining plots

Save each, then combine with **cowplot**.  

```{r}
plot1 <- plot2 <- linelist %>% 
  ggplot(
    mapping = aes(
      x = date_report),
    binwidth = 7)+
  geom_histogram()+
  theme_minimal()+
  scale_y_continuous(expand = c(0,0))+
  scale_x_date(
    expand = c(0,0),
    limits = c(as.Date("2020-03-01"), max(linelist$date_report, na.rm=T)),
    date_breaks = "months",
    labels = scales::label_date_short())

plot1


plot2 <- linelist %>% 
  group_by(week = floor_date(date_report, "week")) %>% 
  summarise(ci = list(mean_cl_normal(age) %>% rename(mean=y, lwr=ymin, upr=ymax))) %>% 
  unnest() %>%  
  ggplot(
    mapping = aes(
      x = week,
      y = mean,
      ymin = lwr,
      ymax = upr))+
  geom_ribbon(alpha = 0.5, fill = "green", color = "green")+
  geom_line(size = 2, color = "darkgreen")+
  scale_x_date(
    limits = c(as.Date("2020-03-01"), max(linelist$date_report, na.rm=T)),
    expand = c(0,0),
    date_breaks = "months",
    labels = scales::label_date_short()
  )+
  coord_cartesian(ylim = c(30, 60))+
  theme_minimal()+
  labs(
    y = "Weekly mean age (95%CI)",
    x = "Month")

plot2


cowplot::plot_grid(plot1, plot2, rel_heights = c(1, 1), ncol = 1, align = "hv")


```


# Dual-axis plots

Save, then combine with **cowplot**.  

```{r}
plot1 <- ggplot(
  data = linelist,
  mapping = aes(
    x = date_report)) + 
geom_histogram(color = "grey", alpha = 0.5)+
theme_cowplot()+
scale_x_date(
 limits = c(
   as.Date("2020-03-01"),
   max(linelist$date_report, na.rm=T)),  
  date_breaks = "months",
  labels = scales::label_date_short(),
  expand = c(0,0),
  name = "")+
scale_y_continuous(
  expand = c(0,0),
  name = "Weekly reported case incidence")

plot1
```


```{r}
plot2 <- linelist %>% 
  group_by(epiweek) %>% 
  summarise(CFR = sum(died == "Yes", na.rm=T) / n() ) %>% 
ggplot(
  mapping = aes(
    x = epiweek,
    y = CFR))+
    geom_line(size = 2, color = "orange")+
  theme_cowplot()+
  scale_y_continuous(
    position = "right",
    limits = c(0, 0.2),
    expand = c(0,0),
    name = "Weekly CFR")+
  scale_x_date(
    limits = c(as.Date("2020-03-01"), max(linelist$date_report, na.rm=T)),  
    date_breaks = "months",
    labels = scales::label_date_short(),
    expand = c(0,0),
    name = "Epiweek")+
  theme(
    axis.text.y = element_text(color = "orange", face = "bold"),
    axis.title.y = element_text(color = "orange", face = "bold"))

plot2
```


```{r}
aligned  <- align_plots(plot1, plot2, align = "hv")
combined <- ggdraw(aligned[[1]]) + draw_plot(aligned[[2]])
combined
```



# Interactive plots  

It is relatively easy to use `ggplotly()` from **plotly**  

```{r echo = T, eval=T}
bar_plot <- ggplot(
  data = linelist,
  mapping = aes(x = eth_race, fill = died)) +             
geom_bar() + 
theme(axis.text.x = element_text(angle=30))
```

```{r echo = T, eval=T}
# Provide the name of a defined plot into ggplotly()  
plotly::ggplotly(bar_plot)
```



